Computing likelihoods under the SMC for a general class of ARGs.

### 2023-12-11 paper.tex
update introduction

1/ ARGs are awesome, or general intro about approximations

model is a mathematical simplification of reality that aims to capture essential aspects 
about what we observe.
For example, the coalescent, ...
Formal way of recording what we observe, results in data structure 
with a certain level of precision.
For example, ARG, ...
The structure capturing the details of these intricately 
interwoven paths of inheritance is referred to as an ancestral recombination graph (ARG)

For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history  
where all events are uniquely positioned both in time
and along the genome. [However see Wong et al. 2023].

Reconstructing these genealogies with certainty about the assumed level of detail 
is impossible.
As such, in the past the data structure capturing these inheritance paths 
or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
variable to be integrated out during statistical inference 
\citep{kuhner_2000, griffiths_ancestral_1996}.\\

% 2/ approximations have driven the rate of progress in inference
Classical approach Hudson, Markovian backwards in time.
McVean and Cardin, SMC, when described left-to-right also Markovian 
in that sense. By sacrificing precision on the our ability to capture long 
range linkage information gain tractability.

SMC functioned as the engine of many powerful inference methods 
if the continuous state space for the Markov chain 
(consisting of all possible genealogies) is approximated by a moderately 
sized finite set—typically by enumerating tree topologies and/or discretizing 
time—then inference can be performed efficiently using well-known algorithms 
for hidden Markov models (HMMs).
ARGweaver uses these same insights to develop the first real milestone in ARG inference.\\

% 3/ Recent progress in ARGs through approximate structures

@@

% heuristic methods guided by coalescent principles

% what is an approximate structure: implicit recombination, polytomies or not, ...

@@

% 4/ Big problem: how do you connect such approximate structures each 
% with different levels of detail to models?
These recent developments have also meant however that
we are now using the term ARG for a varying range of approximate structures 
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).

% 5/ what we will do

>>>>

\subsection{ARGs are awesome, or general intro about approximations}
% approximation: level of detail

In population genetics, the questions we try to answer are generally 
about the unobserved processes that gave rise to the genetic variation 
we can observe given a set of samples. 
The coalescent with recombination 
Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
the generative model that allows us to understand the genealogy 
underlying a particular sampleset. 
Datastructure that captures it: ARG.
For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history  
where all events are uniquely positioned both in time
and along the genome. [However see Wong et al. 2023]. 
% say something about the difficulty of precise ARG inference (see more scalable ARG inference later on).

Although ARG, ...,
something about the level of detail.

\subsection{approximations have driven the rate of progress in inference}

Instead, approximations of the coalescence with recombination (CwR) have  
driven our ability to perform statistical genomic inference. In particular, 
the ability to formulate the inference problem as a hidden Markov model (HMM) 
has proven to be instrumental [references]. 
Two crucial approximations: 2 characteristics: left to right, and discetization 
of state space by emumerating tree topologies and/or discretising time,
then inference can be performed efficiently using well-known HMM algorithms.
To improve the ability to simulate the CwR, McVean and Cardin 
provided an approximation of the CwR that is both Markovian backwards in time 
as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
By sacrificing precision on the our ability to capture long range linkage 
information, the sequentially Markovian coalescent (SMC) 
became the engine of many powerful inference methods [references]. 
Later improvements, SMC', closed the gap between the SMC and the CwR even 
further.
A second key approximation is not a generative model, instead the  
copying model \citep{li_2003_modeling} was specifically designed to 
generate CwR-like patterns. This heuristic approach has been successfully 
applied for demographic inference [and other things: ADD] 
\citep{sheehan_2013_estimating, steinrucken_2018}.

Both approaches have further formed the basis of the recent breakthroughs in 
scalable ARG inference methods [references].

ARGweaver uses 
sample an ARG from the posterior distribution given the observed variational 
data. Although require limiting the set of possible node times to be compatible 
with HMM, ARGs they produce are fully detailed in the sense that recombinations 
are explitly inferred. 
All other methods whether partially relying on
using SMC [ARGneedle] or Li and Stephens [tsinfer, relate] or not [Espallier, ADD] 
are heuristic in nature and % keep models/method separate from level of detail!!

\subsection{Recent progress in ARGs through approximate structures}

@@

% heuristic methods guided by coalescent principles (Li and Stephens)

@@

Although all different, all these methods share the fact that the ARGs they infer 
are approximate structures.
% what is an approximate structure: implicit recombination, polytomies or not, ...
They lack the full detail: implicit recombination, ...
In particular, recombination is dealt with implicitly. Recombination events are 
not directly inferred and thus not dated. Tree topology changes from one local 
tree to the next without explicit inference of the one or more recombination 
events required to explain the inferred tree transition.

Note that approximate structure need not be an issue, the fully precise and detailed 
ARG can never be known. Instead, the main question is how to connect these 
approximate structures to the existing generative models.

\subsection{The Big problem: how do you connect such approximate structures to a generative model}

These recent developments have also meant however that
we are now using the term ARG for a varying range of approximate structures 
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).

\subsection{what we will do}
